---
title: "Decision Trees"
author: "INSOFE Lab Activity on Decision Trees"
date: "23 July 2017"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# C5.0 Trees

**NOTE** Before starting this assignment please remember to clear your environment, you can do that by running the following code chunk

```{r}

rm(list=ls())

```

## Goal

* The goal of this activity is to predict wether a patient has liver disease or not based on various patient related attributes


## Agenda 

* Get the data

* Data Pre-processing

* Build a model

* Predictions

* Communication


## Reading & Understanding the Data

### Read the Data

Make sure the dataset is located in your current working directory, or else you can change your working directory using the "setwd()" function.

```{r}

setwd(choose.dir())
dis_data<-read.csv(file = "ilpd_data.csv",header = T)

```

### Understand the data

* Use the str(), summary(), head() and tail() functions to get the dimensions and types of attributes in the dataset

* The dataset has 582 observations and 11 variables

- __The variable descriptions are given below:__

1 - age	 : Age of the patient

2 - gender : Gender of the patient

3 - TB : Total Bilirubin content

4 - DB : Direct Bilirubin content

5 - alk_phos : Alkaline Phosphotase content

6 - alamine : Alamine Aminotransferase content

7 - aspartate : Aspartate Aminotransferase content

8 - TP : Total Protiens content

9 - albumin : 	Albumin content

10 - A/G : Ratio of Albumin and Globulin 

11 - Disease : Whether the patient has liver disease or not 


```{r}

summary(dis_data)
str(dis_data)
head(dis_data)
tail(dis_data)

```

```{r}



```


## Data Pre-processing

### Verify Data Integrity

* Verify if the dataset has missing values

```{r}

sum(is.na(dis_data))

```

* Verify the data types assigned to the variables in the dataset

```{r}

str(dis_data)

```

### Split the Data into train and test sets

* Use stratified sampling to split the data into train/test sets (70/30)

* Use the createDataPartition() function from the caret package to do stratified sampling

```{r}

library(caret)
set.seed(2482)
trainrows<-createDataPartition(y = dis_data$disease,p = 0.7,list = F)
traindata<-dis_data[trainrows,]
testdata<-dis_data[-trainrows,]

```

### Impute the missing values

* Impute missing values using knnImputation() function in both the train and test datasets

```{r}

library(DMwR)
train_imp<-knnImputation(data = traindata,k = 10,scale = T)
test_imp<-knnImputation(data = testdata,k = 10,scale = T)

```

## Build a  Decision Tree

### Model the tree

* Use Quinlan's C5.0 decision tree algorithm implementation from the C50 package to build your decision tree

```{r}

library(C50)
tree_c5<-C5.0(train_imp$disease~., train_imp)

```

* Build a rules based tree

```{r}

rules_c5<-C5.0(train_imp$disease~., train_imp, rules=T)

```

### Variable Importance in trees

* Find the importance of each variable in the dataset

```{r}

C5imp(tree_c5, metric = "usage")

```

### Rules from trees

* Understand the summary of the returned c5.0 rules based on the decision tree model


```{r}

summary(rules_c5)

```


### Plotting the tree

* Call the plot function on the tree object to visualize the tree

```{r, fig.width= 35, fig.height=15}

plot(tree_c5)

```


## Evaluating the model

### Predictions on the test data

* Evaluate the decision tree using the standard error metrics on test data

```{r}

pred<-predict(object = tree_c5, test_imp)

```

* Report error metrics for classification on test data

```{r}

library(caret)
confusionMatrix(data = pred,reference = test_imp$disease,positive = "yes")

```

# CART Trees

**NOTE** Before starting this assignment please remember to clear your environment, you can do that by running the following code chunk

```{r}

rm(list=ls())

```

* The classification and regression trees use gini index in place of the gain ratio (based on information gain) used by the ID3 based algorithms, such as c4.5 and c5.0

## Goal

* The goal of this activity is to predict the heating load of a residential building, if the building parameters are given

* Hence, in the future architects would be able to build more energy efficient buildings as they can optimize the building parameters to reduce the heating load

## Agenda 

* Get the data

* Data Pre-processing

* Build a model

* Predictions

* Communication


## Reading & Understanding the Data

### Read the Data

* Make sure the dataset is located in your current working directory, or else you can change your working directory using the "setwd()" function.

```{r}

setwd(choose.dir())
eng_data<-read.csv(file = "building_energy.csv",header = T)

```

### Understand the data

* Use the str(), summary(), head() and tail() functions to get the dimensions and types of attributes in the dataset

* The dataset has 768 observations and 9 variables

```{r}

summary(eng_data)
str(eng_data)

```

* The variable names are self explanatory, for further information visit http://www.sciencedirect.com/science/article/pii/S037877881200151X

## Data Pre-processing

### Verify Data Integrity

* Verify if the dataset has missing values

```{r}

sum(is.na(eng_data))

```

* Verify the data types assigned to the variables in the dataset

```{r}

# Enter answer here

str(eng_data)

```

### Split the Data

* Split the data into train/test sets (70/30)

```{r}

library(caret)
set.seed(3466)
engrows<-createDataPartition(y = eng_data$heating_load,p = 0.7,list = F)
eng_train<-eng_data[engrows,]
eng_test<-eng_data[-engrows,]

```

## Build a Regression Tree

### Model the tree

* Use the rpart package to build a cart tree to predict the heating load

```{r}

library(rpart)
eng_tree<-rpart(formula = eng_train$heating_load~.,data = eng_train)
printcp(eng_tree)

```

### Tree Explicability

* Print the variable importance

```{r}

eng_tree$variable.importance

```

* Plot the regression tree

```{r, fig.width=8, fig.height=5}

library(rpart.plot)
rpart.plot(eng_tree)

```

## Evaluation on Test Data

* Report error metrics on the test data

```{r}

preds<-predict(object = eng_tree,eng_test)

```

```{r}

library(DMwR)
regr.eval(trues = eng_test$heating_load,preds = preds)

```




















